---
title: "Statistical inference"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!require("ggplot2")) { install.packages('ggplot2', repos = "http://cran.us.r-project.org"); library("ggplot2") }
```

## Confidence Intervals

We can integrate the standard normal to find the following rules of thumb:

- 95% of the density is within about 2 $\sigma$
- 68% is within about 1 $\sigma$

```{r fig.height = 3, echo=FALSE}
df1 <- data.frame(x = rnorm(1000))
p1 <- ggplot(df1, aes(x)) + theme(legend.position="none") + xlim(-4, 4)
p1 <- p1 + geom_histogram(aes(y=after_stat(density)), binwidth=0.2, na.rm=TRUE)
p1 <- p1 + stat_function(fun=dnorm, args=list(mean=0, sd=1), colour="red")
p1 <- p1 + stat_function(fun=dnorm, xlim = c(-1, 1), geom = "area")
vline_data = data.frame(x_int = c(-2, -1, 1, 2))
p1 + geom_vline(aes(xintercept=x_int, colour="red"), vline_data)
```

Another way of looking at it is that we have 95% confidence that the population mean, $\mu$, of some statistic falls inside the interval:

$$
\bar{X}_n - 2 \sigma/\sqrt{n} \leq \mu \leq \bar{X}_n + 2 \sigma/\sqrt{n}
$$

Or, in general,

$$
P(\mu \in \bar{X}_n \pm z_{(1-\alpha/2)} \sigma/\sqrt{n}) = 100(1-\alpha)\%,
$$

$z_{(1-\alpha/2)}$ being the corresponding standard normal quantile.

***

**example**

```{r echo=FALSE}
#' n: # number of rexp samples for each mean
#' k: # number of means
generate_means <- function(n, years, prob) {
  #' collect k means, each with n samples, in an array
  sample_means = NULL
  for (i in 1:years) sample_means = c(sample_means, mean(rbinom(n, size=1, prob=prob)))
  means_df <- data.frame(means = sample_means)
  means_df
}
```

```{r echo=FALSE}
#' https://bookdown.org/yihui/rmarkdown-cookbook/font-color.html
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
```

Let's get away from normals for a second and talk about a different sort of process. Every year an inquiry is made into whether the good citizens of Zilcholand would like to run juice through the water lines instead of water. A yes or no from $N = 20000$ people.

A single referendum might look like this:

```{r fig.height = 3, echo = FALSE}
N <- 20000
years <- 100
df3 <- data.frame(vote = factor(rbinom(N, size=1, prob=0.4)))
df4 <- generate_means(N, years, 0.4)
X_n <- mean(df4$means)
low  <- function(quant) { X_n - qnorm(quant)*sd(df4$means)/sqrt(length(df4$means)) }
high <- function(quant) { X_n + qnorm(quant)*sd(df4$means)/sqrt(length(df4$means)) }
vline_data <- function(quant) { data.frame(x_int = c(low(quant), X_n, high(quant))) }
```

```{r echo=FALSE}
p3 <- ggplot(df3, aes(vote))+ geom_bar()
p3
```

If we take the average every time, after `r years` years we get the mean response, $\bar{X}_n =`r X_n`$, and we can establish with `r colorize("95% confidence", "red")` that the population mean, $\mu$,

$$
`r low(0.95)` \leq \mu \leq `r high(0.95)`
$$

```{r fig.height = 3, echo = FALSE}
p4 <- ggplot(df4, aes(means)) + theme(legend.position="none")
p4 <- p4 + geom_histogram(aes(y=after_stat(density)), binwidth=0.001, na.rm=TRUE)
p4 <- p4 + geom_vline(aes(xintercept=x_int, colour="red"), vline_data(0.95))
p4 + geom_vline(aes(xintercept=x_int, colour="blue"), vline_data(0.99))
```

or with `r colorize("99% confidence", "blue")`:

$$
`r low(0.99)` \leq \mu \leq `r high(0.99)`
$$

***

[<- Statistical distributions](./stat_dist.html)

***

[Orbits!](../index.html)
