---
title: "Orbits again!"
date: "2023-07-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(r2d3)
```

[Last time](./orbits.html) we solved the motion of two bodies numerically from an *inertial* reference frame.

Now we introduce the notion of the center of mass, or *barycenter*, of the system, which is just a sort of weighted average of the positions.

$$
{\bf R}_\mathrm{COG} = \frac{m_0 {\bf r}_0 + m_1 {\bf r}_1}{m_0 + m_1}
$$

It follows that the velocity and acceleration are:

$$
\begin{align}
{\bf \dot{R}}_\mathrm{COG} &= \frac{m_0 {\bf \dot{r}}_0 + m_1 {\bf \dot{r}}_1}{m_0 + m_1} \\
{\bf \ddot{R}}_\mathrm{COG} &= \frac{m_0 {\bf \ddot{r}}_0 + m_1 {\bf \ddot{r}}_1}{m_0 + m_1} \\
\end{align}
$$

Since the only force we consider here is the gravitational attraction between the masses, we found previously that ${\bf F}_0 = -{\bf F}_1$, so substituting ${\bf F}_i = m {\bf \ddot{r}}_i$:

$$
m_0 {\bf \ddot{r}}_0 + m_1 {\bf \ddot{r}}_1 = 0
$$

So we see that ${\bf \ddot{R}}_\mathrm{COG} = 0$: the barycenter is not accelerating, and is therefore an inertial reference frame.

***

**Example**

Our previous example written more compactly:

```{r}
df <- data.frame(m = c(1.0e26, 5.0e25), x = c(100, 150), y = c(120, 50))
```

```{r echo=FALSE, fig.height=2, out.width="50%"}
r2d3(script="../scripts/rcog.js", data=df,  options=list(cog = "true", lines = "true", text = "true", masses = "true"))
```

***

The barycenter lies on the line between the masses and a little closer to the larger of the two, $m_0$.

## Links

- [Index](./index.html)
- [Orbits!](./orbits.html)
