---
title: "Relative motion"
author: "Emilio Miralles"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(r2d3)
```

We now use the approximate masses of the sun and earth system.

```{r}
m_sun <- 2.0e30
m_earth <- 6.0e24
df <- data.frame(m = c(m_sun, m_earth), x = c(100, 150), y = c(120, 50))
```

Whoops, the earth disappears!

```{r echo=FALSE, fig.height=2, out.width="50%"}
r2d3(script="./scripts/rcog.js", data=df, options=list(relative = "true", masses = "true"))
```

The sizes on the figure correspond to the masses, not the dimensions; we are treating these as point particles with no internal structure. This is just to say that the sun is massively more... massive than the earth, so we want to see how that works out for the physics of their motion.

Lets take a moment to review our previous results. We had a state vector of 12 quantities, $S = [{\bf r}_0, {\bf r}_1, {\bf \dot{r}}_0, {\bf \dot{r}}_1]$, and a system of two differential equations describing the motion of the bodies:

$$
\begin{align}
\ddot{\bf r}_0 &= -G m_1 \frac{{\bf r}_0 - {\bf r}_1}{r^3} \\
\ddot{\bf r}_1 &= -G m_0 \frac{{\bf r}_1 - {\bf r}_0}{r^3} \\
\end{align}
$$

Note the signs chosen to show the similar form of the equations.

Now taking ${\bf r}$ as the position of the earth, $m_1$ around the sun, $m_0$, we have equations for its position, velocity, and acceleration:

$$
{\bf r} = {\bf r}_1 - {\bf r}_0 \\
\dot{\bf r} = \dot{\bf r}_1 - \dot{\bf r}_0 \\
\ddot{\bf r} = \ddot{\bf r}_1 - \ddot{\bf r}_0
$$

Substituting the accelerations derived from Newton's [law of gravitation](https://en.wikipedia.org/wiki/Newton%27s_law_of_universal_gravitation), we develop the equations of motion.

$$
\begin{align}
\ddot{\bf r} &= \ddot{\bf r}_1 - \ddot{\bf r}_0 \\
\ddot{\bf r} &= -G m_0 \frac{\bf r}{r^3} - G m_1 \frac{\bf r}{r^3} \\
\ddot{\bf r} &= -G(m_0 + m_1) \frac{\bf r}{r^3}
\end{align}
$$

we now have a single equation, with only six quantities, $S = [{\bf r}, {\bf \dot{r}}]$.

## Scalar form

Writing the *standard gravitational parameter*, $\mu = G(m_0 + m_1)$,

$$
\ddot{\bf r} = -\mu \frac{\bf r}{r^3} = -\mu \frac{\hat{\bf r}}{r^2}
$$

So the acceleration is along the path between the two and decreases as the square of the distance. The equations are proportional to some negative constant, signalling an attractive force relating to a characteristic mass.

We want to recast this vector equation into some scalar form and see if we can tease out some of the properties of the motion.
