---
title: "Central limit theorem"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

The [law of large numbers][1] shows that the average of many samples will approach the expected value as the number of samples goes to infinity:

$$
\lim_{n \to \infty} \sum_{i=1}^n \frac{X_i}{n} = E(X).
$$

For [independent and identically distributed variables][2], the [central limit theorem][3] shows that as your sample size increases, the distribution of averages approaches a [standard normal distribution][4].

The normalization moves the values so that that $\mu = 0$ and $\sigma = 1$.

$$
X = \frac{X' - \mu}{\sigma/\sqrt{n}}.
$$

### Exponential

We explore the exponential distribution via simulation and compare our sample statistics with the theoretical statistics of the distribution.

An exponential distribution can describe events that happen at a constant average rate, independently from each other. It has a density function,

$$
f(x) = \lambda e^{- \lambda x},
$$

for $x \geq 0$. We define our rate parameter ($\lambda$), calculate the theoretical population mean ($\mu$), standard deviation ($\sigma$), and variance ($\sigma^2$):

```{r}
lambda = 0.2 # rate, mu = 5, sd = 5, var = 25
pop_mean = 1/lambda
pop_sd = 1/lambda
pop_var = pop_sd^2
```

```{r echo=FALSE}
#' n: # number of rexp samples for each mean
#' k: # number of means
generate_means <- function(n, k, lambda) {
  #' collect k means, each with n samples, in an array
  sample_means = NULL
  for (i in 1:k) sample_means = c(sample_means, mean(rexp(n, lambda)))
  means_df <- data.frame(means = sample_means)
  means_df
}
```

```{r}
n <- 40
k <- 1000
df1 <- generate_means(n, k, lambda)
```

We take $`r k`$ means of exponential samples each with $n = `r n`$.

## Sample Mean versus Theoretical Mean

The mean of our sampled means is $`r round(mean(df1$means),3)`$, close to the theoretical value of $`r pop_mean`$ represented by the red line on the boxplot.

```{r fig.height = 3, echo=FALSE}
p1 <- ggplot(df1, aes(means)) + theme(legend.position="none")
p1 + geom_boxplot() + geom_vline(aes(xintercept = pop_mean, colour = "red"))
```

## Sample Variance versus Theoretical Variance

If we assume $n = `r n`$ is large enough, the distribution is close to normal with mean $\mu$ and variance $\sigma^2/n$, where $\sigma^2$ is the variance of the exponential distribution we were originally sampling from. The variance $\sigma^2/n$ is referred to as the *standard error*, and will be our theoretical variance.

| exponential variance ($\sigma^2$) | theoretical variance ($\sigma^2/n$) | sample variance         |
|:---------------------------------:|:-----------------------------------:|:-----------------------:|
| $`r pop_var`$                     | $`r pop_var/n`$                     | $`r var(df1$means)`$ |

## Distribution

Since we are sampling from the `rexp` function, we know the data is independent and identically distributed. A large number of `rexp` samples simply follow the exponential curve. We are interested in the means of these samples, so the central limit theorem applies.

```{r fig.height = 3, echo = FALSE}
sample <- data.frame(data=rexp(k, lambda))
p2 <- ggplot(sample, aes(data))
p2 <- p2 + geom_histogram(aes(y=after_stat(density)), binwidth=1.5)
p2 + stat_function(fun=dexp, args=list(rate = lambda), colour="red", xlim=c(0, max(sample$data)))
```

While the underlying samples are exponential, the distribution of means looks normal (dashed red line), centered around the theoretical population mean $\mu$ (the vertical red line), with standard deviation $\sigma/\sqrt{n}$.

```{r fig.height = 3, echo=FALSE}
p3 <- ggplot(df1, aes(means)) + theme(legend.position="none")
p3 <- p3 + geom_histogram(aes(y=after_stat(density)), binwidth=0.2)
p3 <- p3 + stat_function(fun=dnorm, args=list(mean=pop_mean, sd=pop_sd/sqrt(n)), colour="red")
p3 + geom_vline(aes(xintercept=pop_mean, colour="red"))
```

Any individual sample of the exponential distribution can have a mean far from the population mean, as seen for cumulative means with small $k$ on the left. As we average over more sample means, we get closer to the theoretical population mean $\mu = `r pop_mean`$, the horizontal red line.

```{r fig.height = 3, echo=FALSE}
cumulative <- data.frame(means=cumsum(df1$means) / (1 : k))
p4 <- ggplot(cumulative, aes(x=1:k, y=means)) + theme(legend.position="none")
p4 + geom_line() + geom_hline(aes(yintercept=pop_mean, colour="red"))
```

***

- [next: Statistical inference](./stat_inf.html)
- [prev: Statistical distributions](./stat_dist.html)

[1]:(https://en.wikipedia.org/wiki/Law_of_large_numbers)
[2]:(https://en.wikipedia.org/wiki/Independent_and_identically_distributed_random_variables)
[3]:(https://en.wikipedia.org/wiki/Central_limit_theorem)
[4]:(https://en.wikipedia.org/wiki/Normal_distribution)

***

[Orbits!](./index.html)
